Asthma Cohort Descriptive Epidemiology

### 2017-12-06 r_scripts/analysis/asthma_cohort_descriptive_epi.Rmd
started asthma cohort exploration.

---
title: Distributed lag effects of smoke on cardiovascular disease in persons over 65
author: "Ryan Gan"
date: "2017-11-20"
output: html_document
---

@@

Extensive look at possible distributed lag relationships for Cardiovascular endpoints in the over 65 age group.

Packages used: tidyverse, survival. Setting markdown options.

@@

Importing the time-stratified case-crossover dataframes for the cardiovascular endpoints.

```{r health data import}
# path for health dataframe
health_path <- "../../data/health/"
# importing time-stratified case-crossover dataframe
casecross_files <- list.files(path=health_path, pattern="*casecross*")

# reorder file names in way I'd like it to be analyzed and only importing cvd
# also ditching broken arm
casecross_order <- casecross_files[c(7,2,5,8,9,10)]

# extract the name of the dataframe after study type
file_names <- stringr::str_extract(casecross_order, 
  pattern="(?<=casecross_)\\w+")

# read in all the files in to a list
cvd_over65_list <- casecross_order %>%  
  purrr::map(~read_csv(paste0(health_path, .))) %>% 
  # filter to over 65 group
  purrr::map(~filter(., age >= 65))

# name dataframes
names(cvd_over65_list) <- file_names

# name list elements in order I'd like to present in document
outcome_names <- c("Cardiovascular Disease", "Arrhythmia", 
  "Cerebrovascular Disease", "Heart Failure", "Ischemic Heart Disease", 
  "Myocardial Infarction")
```

Creating a fit function to find the degrees of freedom for the pm spline

```{r custom distributed lag fit function}
# fit function
dl_fit <- function(data, pm_df, temp_df){
  # i want to remove this part eventually ------
  # limit to complete cases
  complete_data <- data %>% 
    # filter to complete case for smoke
    filter(!is.na(geo_smk_pm_lag7_zip))

  # output matrix of gwr values
  pm_matrix <- complete_data %>% 
    select(contains("geo_smk_pm")) %>%
    # remove anything with m in it; will go back in original code and not create 
    # this to begin with
    select(-contains("_m")) %>% 
    # divide exposure values by 10 units to interpret on 10 ug/m^3 scale
    mutate_all(funs(./10)) %>% 
    # convert to matrix
    as.matrix()

  # output temp matrix
  temp_matrix <- complete_data %>% 
    select(contains("wrf_temp")) %>%
    # remove anything with m in it; will go back in original code and not create 
    # this to begin with
    select(-contains("_m")) %>% 
    # convert to matrix
    as.matrix()
  # end part to remove------

    # create temp basis
    temp_b <- splines::ns(0:(ncol(temp_matrix)-1), df=temp_df, intercept=T)
    temp_basis <- temp_matrix %*% temp_b

      # create list of lengh pm_df
      mod_aic <- sapply(pm_df, function(x){
        pm_b <- splines::ns(0:(ncol(pm_matrix)-1), df=x, intercept=T)
        # create pm basis
        pm_basis <- pm_matrix %*% pm_b
        # run model
        mod <- clogit(outcome ~ pm_basis + temp_basis + strata(personkey),
          data = complete_data)
        aic <- AIC(mod)
        return(aic)
      }) # end sapply

    # find minimum aic
    pm_df_aic <- data.frame(pm_df, mod_aic)
    min_df_aic <- subset(pm_df_aic, mod_aic == min(mod_aic))
    return(min_df_aic)
  } # end fit option
```

```{r custom distributed lag function}
# custom distributed lag function 
distributed_lag <- function(data, pm_df, temp_df, cumulative=T){
  # i want to remove this part eventually ------
  # limit to complete cases
  complete_data <- data %>% 
    # filter to complete case for smoke
    filter(!is.na(geo_smk_pm_lag7_zip))

  # output matrix of gwr values
  pm_matrix <- complete_data %>% 
    select(contains("geo_smk_pm")) %>%
    # remove anything with m in it; will go back in original code and not create 
    # this to begin with
    select(-contains("_m")) %>% 
    # divide exposure values by 10 units to interpret on 10 ug/m^3 scale
    mutate_all(funs(./10)) %>% 
    # convert to matrix
    as.matrix()

  # output temp matrix
  temp_matrix <- complete_data %>% 
    select(contains("wrf_temp")) %>%
    # remove anything with m in it; will go back in original code and not create 
    # this to begin with
    select(-contains("_m")) %>% 
    # convert to matrix
    as.matrix()
  # end part to remove------

  # calculation of basis ----
  # define basis using natural spline function from "splines"" package
  pm_b <- splines::ns(0:(ncol(pm_matrix)-1), df=pm_df, intercept=T)
  # create pm basis
  pm_basis <- pm_matrix %*% pm_b
  # create temp basis
  temp_b <- splines::ns(0:(ncol(temp_matrix)-1), df=temp_df, intercept=T)
  temp_basis <- temp_matrix %*% temp_b
  # fit model with basis ----
  mod <- clogit(outcome ~ pm_basis + temp_basis + strata(personkey),
    data = complete_data)

  # calculate estimates ----
  # output pm basis parameters
  dl_parms <- broom::tidy(mod) %>% 
    filter(stringr::str_detect(term, "pm")) %>% 
    select(estimate) %>% 
    as_vector()
  # estimate distributed lag values for each day
  estimate <-  pm_b %*% dl_parms
  # covariance matrix for knots 
  # fix the matrix
  cov_matrix <- as.matrix(vcov(mod))[1:pm_df,1:pm_df]
  # estimate variance of spline
  variance <- pm_b %*% cov_matrix %*% t(pm_b)
  # estimate standard error
  se <- sqrt(diag(dl_var))
  # calculate lower and upper bound
  lower95 <- estimate+(se*qnorm(1-0.975))
  upper95 <- estimate+(se*qnorm(0.975))
  # time variable
  time <- ((rep(1:length(estimate))-1))

    if(cumulative==T) {
      type <- rep("cumulative", times = length(estimate))
      # sequential cumulative estimate
      cumulative_estimate <- sapply(seq_along(estimate), function(x){
        sum(estimate[1:x])
      })
      # stderr cumulative effect
      cumulative_se <- sapply(seq_along(estimate), function(x){
        sqrt(sum(dl_var[1:x,1:x]))
      })
      # cumulative 95CI
      lower95 <- cumulative_estimate+(cumulative_se*qnorm(1-0.975))
      upper95 <- cumulative_estimate+(cumulative_se*qnorm(0.975))
      # return dataframe
      return_estimate <- data.frame(type, time, exp(cumulative_estimate), 
                                    exp(lower95), exp(upper95)) 
      # assign column names
      colnames(return_estimate) <- c("type","time","estimate","lower95","upper95")

      return(return_estimate)
    } else { # if cumulative is not true, or false, return distributed lag est
    type <- rep("lag", times = length(estimate))
    # return dataframe
    return_estimate <- data.frame(type, time, exp(estimate), 
                                  exp(lower95), exp(upper95)) 
    # assign column names
    colnames(return_estimate) <- c("type","time","estimate","lower95","upper95") 
    # return estimate  
    return(return_estimate)
    } # end of if
} 
```

## Finding smoke distributed lag fit based on AIC

This chunk of code uses the dl_fit function I created to find the degrees of freedom for the natural spline that has the lowest AIC for degrees of freem from 2 to 5 for the smoke lag function. I set temperature degrees of freedom to 2.

```{r smk aic fit}
fit_df <- cvd_over65_list %>% 
  purrr::map_df(~dl_fit(., pm_df = c(2:5), temp_df =2)) %>% 
  cbind(names(cvd_over65_list), .)

# add degrees of freedom as a slot to the outcomes list
degfreedom_list <- as.list(fit_df$pm_df)
```

## Distributed Lag Cumulative Effect

Calculating the cumulative effect sequentially (0, 0+1, 0+1+2,....)

```{r cumulative effect smooth}
cumulative_cvd <- purrr::map2(.x = cvd_over65_list, .y = degfreedom_list, 
  ~distributed_lag(data = .x, pm_df = .y, temp_df = 2, cumulative = T)) %>% 
  plyr::rbind.fill() %>% 
  bind_cols(., data_frame(rep(outcome_names, each=8))) %>% 
  rename(outcome_names = `rep(outcome_names, each = 8)`) %>% 
  select(outcome_names, type, time, estimate, lower95, upper95) %>% 
  # preserve order of names
  mutate(outcome_names = parse_factor(outcome_names, levels = outcome_names))
```

## Heart failure in over 65 group

```{r subset data}
hf65 <- outcome_df_list[[4]] %>% 
  filter(age_ind == 2)

cumulative <- distributed_lag(hf65, pm_df = 2, temp_df = 2, cumulative = T)
cumulative

#plot 
cumulative_plot <- ggplot(cumulative, aes(x=time, y=estimate)) +
  geom_line(colour = "#046C9A", size = 1) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), 
              fill = "#ABDDDE", alpha = 0.5) + 
  scale_y_continuous(limits = c(0.9, 1.25)) +
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  ylab(expression("Odd ratio for a 10 ug/m^3 increase smoke PM"[2.5])) +
  xlab("Lagged Days") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    strip.text = element_text(size = 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(colour=NA, fill=NA))

cumulative_plot

dl <- distributed_lag(hf65, pm_df = 3, temp_df = 2, cumulative = F)
dl

dl_plot <- ggplot(dl, aes(x=time, y=estimate)) +
  geom_line(colour = "#046C9A", size = 1) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), 
              fill = "#ABDDDE", alpha = 0.5) + 
  scale_y_continuous(limits = c(0.9, 1.25)) +
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  ylab(expression("Odd ratio for a 10 ug/m^3 increase smoke PM"[2.5])) +
  xlab("Lagged Days") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    strip.text = element_text(size = 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(colour=NA, fill=NA))

dl_plot
```

## Myocardial infarction in over 65 group

```{r mi data}
mi65 <- outcome_df_list[[6]] %>% 
  filter(age_ind == 2)

check <- mi65 %>% filter(personkey == 1782923) %>% 
  select(personkey, outcome, dx1, date, fromdate)

cumulative <- distributed_lag(mi65, pm_df = 2, temp_df = 2, cumulative = T)
cumulative

#plot 
cumulative_plot <- ggplot(cumulative, aes(x=time, y=estimate)) +
  geom_line(colour = "#046C9A", size = 1) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), 
              fill = "#ABDDDE", alpha = 0.5) + 
  scale_y_continuous(limits = c(0.7, 1.2)) +
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  ylab(expression("Odd ratio for a 10 ug/m^3 increase smoke PM"[2.5])) +
  xlab("Lagged Days") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    strip.text = element_text(size = 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(colour=NA, fill=NA))

cumulative_plot

dl <- distributed_lag(mi65, pm_df = 3, temp_df = 2, cumulative = F)
dl

dl_plot <- ggplot(dl, aes(x=time, y=estimate)) +
  geom_line(colour = "#046C9A", size = 1) +
  geom_ribbon(aes(ymin = lower95, ymax = upper95), 
              fill = "#ABDDDE", alpha = 0.5) + 
  scale_y_continuous(limits = c(0.9, 1.25)) +
  scale_x_continuous(breaks = c(seq(0,7, by=1))) +
  geom_hline(yintercept = 1, linetype = 2, colour = "red") +
  ylab(expression("Odd ratio for a 10 ug/m^3 increase smoke PM"[2.5])) +
  xlab("Lagged Days") +
  theme(panel.background = element_rect(fill = 'white', colour = 'black'),
    strip.text = element_text(size = 10),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(colour=NA, fill=NA))

dl_plot

```

>>>>

---
title: "Asthma Cohort Descriptive Epidemiology"
author: "Ryan Gan"
date: "2017-12-06"
output: html_document
---

@@

Descriptive statistics of the assembled asthma cohort. This cohort is made up of all billing records for procedures or pharamacy fills for anyone in the Oregon APAC data that had been billed for a diagnosis of asthma at any visit in 2013.

@@
